---
title: Estimating the sleep need of adolescents using nonlinear mixed effects modelling
output: 
  github_document:
    pandoc_args: --webtex
    dev: png
    toc: true
---


# Introduction
How much sleep is enough sleep? Using changes in lapses during the psychomotor vigilance task (PVT), Van Dongen et al. (2003) estimated that it is close to 8.16h. In their study, participants were split into different sleep restriction conditions and lapses were analysed for each of these conditions across several days. Lapses in a PVT task are when participants miss reacting to a stimulus that they are supposed to. In general, the number of lapses increases as the amount of sleep restriction increases. Van Dongen et al. modelled this alternatively as lapses increasing as the amount of wakefulness during the previous day increase beyond a certain critical value. This was formulated mathematically as follows:

$$lapses = b(excess)^\theta$$

Here, b represents rate of change of lapses per unit change in the nonlinear part of the equation. Excess denotes sleep in excess of the critical waking duration. Theta accomodates nonlineaerity in the relationship. Excess can then be formulated as the difference between Cumulative Wake Time (CWT) across a number of days and the critical waking duration multiplied across the number of days.

$$lapses = b(CWT - critical*day)^\theta$$

Van Dongen's study was based on a population of adults between the ages 21 and 38. In this analysis, I use the same approach but for adolescents between the ages of 15 and 19. The data here is from the Need for sleep studies (Lo et al, 2016; 2017; 2019; 2020). There are four different sleep conditions - 5h, 6.5 h, 8h and 9h across the studies. The baseline sleep for all sleep conditions was 9h. Participants took the neurobehavioral tests three times every day across the sleep restriction days. For the analysis, I need the average lapses across days. These studies also had periods of recovery sleep, either inbetween or at the end. However, only the baseline sleep plus the days of sleep restriction until the first sleep recovery period is sufficient for this analysis. 



```{r setup, include=FALSE}
knitr::opts_chunk$set(knitr::opts_chunk$set(warning = FALSE, message = FALSE), echo = TRUE)

```

# The Dataset

The dataset for this analysis comes from NFS1, NFS2, NFS4 and NFS5. Here is a section of the original dataset. Most of the columns have been omitted for the sake of presentation here.

```{r, message=FALSE, echo=FALSE}

library(tidyverse)
library(rmarkdown)

#load data
data.nfs = read.csv("nfs.csv")
print(head(data.nfs[, 1:8])) #display dataset

```

## Data cleaning

First, I select only the columns that are required for the analysis.

```{r, message = FALSE, echo=FALSE}

#select only the required groups
data.nfs = data.nfs %>% select(subj, day_num, group, lapses = pvt_lapses)
data.nfs = data.nfs %>% na.omit() #remove NAs
print(head(data.nfs)) #view dataset

```

Since participants did the PVT task three times a day, I summarised the results to get the average PVT lapses for each day.

```{r, message=FALSE, echo=FALSE}

#summarize to get mean pvt lapses
data.nfs = data.nfs %>% group_by(subj, group, day_num) %>% 
  summarise(lapses = mean(lapses))
print(head(data.nfs))

```

I imported the TST data.

```{r, echo=FALSE}

nfs.tst <- data.frame() #create an empty dataframe to initialize

#run a loop to read TST data
for(i in c(1, 2, 4, 5)){
  study <- paste("NFS", i, sep = "")
  if(i %in% c(1, 2)){
  nfs.tst.temp <- read.csv(paste(study, ".csv", sep = "")) 
  names(nfs.tst.temp)[2] = "lab"
  nfs.tst.temp <- nfs.tst.temp %>% 
    select(subj = Subject.ID, 
           sleep_time = Total.Sleep.Time..hours., lab) %>%
                        filter(lab != 0) %>% select(-lab)
    if(i == 1){
    day.c <- data.frame(day = rep(1:12, times = nrow(nfs.tst.temp %>% 
                                                       group_by(subj) %>% 
                                                       count())))
    nfs.tst.temp$day <- day.c$day
    }
    else{
    day.c <- data.frame(day = rep(1:15, 
                                  times = nrow(nfs.tst.temp %>% 
                                               group_by(subj) %>% 
                                               count())))
    nfs.tst.temp$day <- day.c$day
    }
  }
  else{
    nfs.tst.temp <- read.csv(paste(study, ".csv", sep = "")) %>% 
      select(subj = subID, sleep_time = Total.sleep.time)
    nfs.tst.temp <- nfs.tst.temp %>% filter(subj != "")

    irreg <- nfs.tst.temp %>% group_by(subj) %>% count() %>% 
      filter(n < 14) %>% select(subj)
    nfs.tst.temp <- nfs.tst.temp %>% filter(!subj %in% irreg$subj)
  
    day.c <- data.frame(day = rep(1:14, times = nrow(nfs.tst.temp %>% 
                                                     group_by(subj) %>% 
                                                     count())))
    nfs.tst.temp$day <- day.c$day
  }
  nfs.tst.temp$study = study
  nfs.tst <- rbind(nfs.tst, nfs.tst.temp)
}

nfs.tst <- na.omit(nfs.tst)

#remove day 1 from NFS1 and subtract 1 from all days in NFS1 
nfs.tst <- nfs.tst %>% filter(ifelse(study == "NFS1", day != 1, day > 0))
nfs.tst <- nfs.tst %>% mutate(day = ifelse(study == "NFS1", day - 1, day))

#fix the timeline in the same way for NFS2
#nfs.tst %>% group_by(subj) %>% filter(study == "NFS2")
nfs.tst <- nfs.tst %>% filter(ifelse(study == "NFS2", day %in% c(3 : 14), day > 0))

#fix the timeline for NFS4 and NFS5
#nfs.tst %>% group_by(subj) %>% filter(study == "NFS4")
nfs.tst <- nfs.tst %>% filter(ifelse(study == "NFS4", day %in% c(2 : 13), day > 0))

#nfs.tst %>% group_by(subj) %>% filter(study == "NFS5")
nfs.tst <- nfs.tst %>% filter(ifelse(study == "NFS5", day %in% c(2 : 13), day > 0))

nfs.tst <- nfs.tst %>% mutate(day = ifelse(study %in% c("NFS4", "NFS5"), day + 1, day))

nfs.tst$sleep_time <- as.numeric(nfs.tst$sleep_time) #convert data to numeric
nfs.tst <- nfs.tst %>% mutate(sleep_time = ifelse(sleep_time > 24, sleep_time/60, sleep_time)) #convert minutes to hours (TST)

#filtering to match participants in both datasets
subj.tst <-  nfs.tst %>% group_by(subj) %>% count() %>% select(subj)
data.nfs <- data.nfs %>% filter(subj %in% subj.tst$subj)
subj.nfs <-  data.nfs %>% group_by(subj) %>% count() %>% select(subj)
nfs.tst <- nfs.tst %>% filter(subj %in% subj.nfs$subj)


#nrow(data.nfs) == nrow(nfs.tst) #check if row count is same
names(nfs.tst)[3] = "day_num" #name day column to match with that on data.nfs dataframe
names(nfs.tst)[2] = "TST" #name sleep_time column to 'TST'

print(head(nfs.tst))
```

Merge sleep time data and NFS dataset

```{r, echo=FALSE}
common_names <- intersect(names(data.nfs), names(nfs.tst))
data.safe <- data.nfs
#data.nfs <- data.safe
data.nfs <- merge(data.nfs, nfs.tst, by = common_names, all.x = TRUE)
#data.nfs %>% filter(study == "NFS1")
print(head(data.nfs))
```

I create a new column that denotes their bed time during the previous night. On baseline days, they have 9h of sleep and their sleep time varies based on their sleep condition on the other days. The day numbers are also made to align across the different studies and start with day 1 being the baseline day.

```{r, echo=FALSE}

#Create a new column 'TBT' indicating sleep time on the previous night:

#NFS 1, nonap_5hx7 - data starts from day 1
#day 1 is after baseline; 9, 10, 11 are after recovery nights
data.nfs = data.nfs %>% 
  mutate(TBT = ifelse(group == "nonap_5hx7" & day_num %in% c(1, 9, 10, 11), 9.0, 0.0)) #assign 9h

#NFS 1, control group
data.nfs = data.nfs %>% 
  mutate(TBT = ifelse(group == "control_9h", 9, TBT))

#Other NFS - data starts from day 3
#day 3 is baseline, 9, 10 are after the intermediate recovery nights
#day 14 is the final recovery night
data.nfs = data.nfs %>% 
  mutate(TBT = ifelse(group != "nonap_5hx7" & day_num %in% c(3, 9, 10, 14), 9.0, TBT))

data.nfs <- data.nfs %>% mutate(day_num = ifelse(group %in% c("nonap_5hx7", "control_9h"), day_num + 2, day_num))#bring the baseline day to 3 for NFS 1

#initialise condition
data.nfs <- data.nfs %>% mutate(condition = 0)

#rename groups to numeric value denoting sleep condition
data.nfs$condition[data.nfs$group == "nonap_5hx7"] = as.numeric(5)
data.nfs$condition[data.nfs$group == "nonap_5h"] = as.numeric(5)
data.nfs$condition[data.nfs$group == "nonap_6pt5h"] = as.numeric(6.5)
data.nfs$condition[data.nfs$group == "nonap_8h"] = as.numeric(8)
data.nfs$condition[data.nfs$group == "control_9h"] = as.numeric(9)
data.nfs$condition[data.nfs$group == "nap60_5h"] = as.numeric(6)
data.nfs$condition[data.nfs$group == "nap90_5h"] = as.numeric(6.5)
data.nfs$condition[data.nfs$group == "nap90_6pt5h"] = as.numeric(8)

data.nfs = data.nfs %>% 
  mutate(TBT = ifelse(TBT == 0.0, as.numeric(condition), TBT))#set TBT for other days using sleep conditions

#align the day numbers between the different NFS together

data.nfs = data.nfs %>% 
  mutate(day_num = day_num - 2)#subtract 2 from all to bring baseline day to 1

print(head(data.nfs))
```

Finally I calculate the cumulative wake duration for each participant

```{r, echo=FALSE}

#Calculate cumulative wake duration

#based on TBT
data.nfs = data.nfs %>% group_by(subj) %>% 
  mutate(TWT_tbt = cumsum(24 - TBT))

#based on TST
data.nfs = data.nfs %>% group_by(subj) %>% 
  mutate(TWT_tst = cumsum(24 - TST))

#Filter only days until the first 6 days (1 baseline and 5 manipulation)
data.nfs = data.nfs %>% filter(day_num <= 6)

#data.allconditions <- data.nfs

data.naponly = data.nfs %>% filter(group %in% c("nap60_5h", "nap90_5h", "nap90_6pt5h"))

data.nonap = data.nfs %>% filter(group != "nap60_5h",
                                group != "nap90_5h",
                                group != "nap90_6pt5h")
data.nonap <- na.omit(data.nonap)

print(head(data.nfs))
```

## Clean data

Summing up, the last baseline day plus the first five sleep manipulation days were used for the analysis from the NFS studies. ‘Lapses’ represents mean lapses on a particular day (day_num). Sleep Time (ST) represents the total sleep time on the previous night. Total Wake Time (TWT) was calculated as the cumulative total wakeful duration across days (24 – ST*day_num). After cleaning, there were 852 observations in total (6 days x 142 subjects).

```{r, echo=FALSE}

print(head(data.nfs))

```

# Visualising the dataset

For this analysis, I have mainly focused on no nap conditions. Hence the visualizations below do not include the conditions that had nap time.

## Boxplot - Lapses vs Sleep Condition

```{r, echo = FALSE, warning=FALSE, message=FALSE, fig.height = 8, fig.width=10, dpi=300}

# Lapses across groups

day_labels <- c('1' = 'day 1',
                   '2' = 'day 2',
                   '3' = 'day 3',
                   '4' = 'day 4',
                   '5' = 'day 5',
                   '6' = 'day 6')

ggplot(data.nonap, aes(x = as.factor(condition), y = lapses)) +
   geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  # geom_jitter(alpha = 0.1) +
   labs(x = "condition", y = "lapses", title = "Lapses vs Sleep condition across 6 days", fill = "Condition") + 
   #stat_summary(fun = mean, geom = "line", size = 1, alpha = 0.8, aes(group=1, color = as.factor(condition))) +
  facet_wrap(~ day_num, labeller = as_labeller(day_labels)) +
  guides(color = FALSE) + theme_minimal()
  

```

As it can be seen, the number of lapses increases across the days as the level of sleep restriction increases. This can be noted more clearly in the following charts



```{r, message=FALSE, fig.align="center", include=FALSE}
### Scatter plot of lapses across days divided by condition
#lapses vs day

ggplot(data.nonap, aes(x = day_num, y = lapses) ) +
  #geom_ribbon() +
  geom_jitter(aes(y = lapses, color = factor(condition)), alpha = 0.3) +
  labs(x = "day", y = "mean lapses", title = " ", color = "condition") +
  facet_wrap(~ as.factor(condition))

```

## Line plot - Lapses vs Day

```{r, echo = FALSE, message=FALSE, fig.height = 6, fig.width=10, dpi=300}

#mean lapses vs day

ggplot(data.nonap, aes(x = day_num, y = lapses, color = as.factor(condition))) +
   stat_summary(fun = mean, geom = "line", size = 1, alpha = 0.5) +
   stat_summary(fun = mean, geom = "point") +
   stat_summary(fun.data = mean_cl_normal, geom = "errorbar", aes(group = as.factor(condition)), width = 0.05) +
   labs(x = "day", y = "mean lapses", title = " ", color = "condition") + theme_minimal()

```

# Estimating Sleep Need
## TBT based estimate
### Summary

I'm using the nlme() function from the nlme library to fit a nonlinear mixed effects model to my data.

```{r, message=FALSE}

library(nlme)
TBT.nonap.lapses <- nlme(lapses ~ b*(TWT_tbt - crit*day_num)^theta,
                   data = data.nonap,
                   fixed = b + theta + crit ~ 1,
                   random = crit ~ 1,
                   groups = ~ subj,
                   start = c(b = 1.4, theta = 0.4, crit = 13),
                   na.action = na.omit
                   ) 

summary(TBT.nonap.lapses)

```

### Estimates and their 95% confidence intervals

```{r, message=FALSE}

intervals(TBT.nonap.lapses)

```

## TST based estimate
### Summary

```{r, message=FALSE}

#SLEEP TIME ESTIMATES - no nap
TST.nonap.lapses <- nlme(lapses ~ b*(TWT_tst - crit*day_num)^theta,
                   data = data.nonap,
                   fixed = b + theta + crit ~ 1,
                   random = crit ~ 1,
                   groups = ~ subj,
                   start = c(b = 1.4, theta = 0.5, crit = 12),
                   na.action = na.omit
                   ) 

summary(TST.nonap.lapses)

```

### Estimates and their 95% confidence intervals

```{r, message=FALSE}

intervals(TST.nonap.lapses)

```

# Visualising the results of TST based estimate
## Dot Plot - Critical Wake Durations across participants

```{r, echo = FALSE, message=FALSE, fig.height = 3, fig.width=5, dpi=300}

#extract critical wake duration
crit <- TST.nonap.lapses$coefficients$random$subj + TST.nonap.lapses$coefficients$fixed[3]
crit <- data.frame(crit)

crit$id <- 1:nrow(crit)

ggplot(crit, aes(x = crit, y = id)) +
  geom_point(alpha = 0.5, color = "darkgreen") +
  geom_vline(xintercept = TST.nonap.lapses$coefficients$fixed[3], color = "navyblue", linetype = 2, size = 0.7) +
  labs(x = "critical waking duration", y = "id")
```

## QQ Plot - Normality of distribution

```{r, echo = FALSE, message=FALSE, fig.height = 3, fig.width=5, dpi=300}

#extract sd of random effects
sdcrit <- intervals(TST.nonap.lapses)
sdcrit <- sdcrit$reStruct$subj$est.

qqnorm(crit$crit, pch = 1, frame = FALSE)
qqline(crit$crit, col = "steelblue", lwd = 2)


```

## Histogram - Normality of distribution

The plot indicates a left skew. Let's check the histogram

```{r, echo = FALSE, message=FALSE, fig.height = 3, fig.width=5, dpi=300}


ggplot(crit, aes(x = crit)) +
  geom_histogram(bins = 50, fill = "black", color = "white", alpha = 0.8, aes(y = ..count..)) 

```

## Residuals Plot

```{r, echo = FALSE, message=FALSE, fig.height = 3, fig.width=5, dpi=300}

plot(TST.nonap.lapses)

```

## Comparison against estimates based on the literature

There are clearly some outliers having oversized influence on the distribution. Let's assume normalty for now and compare the distribution obtained from this estimate with Van Dongen et al.(2003) and Short et al. (2018) estimates for the critical waking duration

```{r, echo = FALSE, message=FALSE, fig.height = 4, fig.width=6, dpi=300}


#plot different estimates
ggplot(crit, aes(x = crit)) +
  #stat_function(fun = dnorm, args = list(15.88, 1.42), color = "red", size = 1, alpha = 0.5) +
  geom_area(stat = "function", fun = dnorm, args = list(mean = TST.nonap.lapses$coefficients$fixed[3], sd = sdcrit), aes(fill = "NFS, age: 15 - 19"), alpha = 0.2) +
  geom_area(stat = "function", fun = dnorm, args = list(mean = 15.82, sd = 3.84), aes(fill = "Van Dongen et al., age: 21 - 38"), alpha = 0.2)+
  geom_area(stat = "function", fun = dnorm, args = list(mean = 14.65, sd = 4.14), aes(fill = "Short et al., age: 15 - 17"),  alpha = 0.2)


```

## Mean lapses vs excess wakefulness

The line represents predicted number of lapses, dots represent the actual number of lapses observed.

```{r, echo = FALSE, message=FALSE, fig.height = 4, fig.width=6, dpi=300}

# #lapses predictions
# data.nonap$plapses = predict(sleep.nonap.lapses, data.nonap)
# data.nonap$plapses[is.nan(data.nonap$plapses)] <- 0.0
# 
# #graphing lapses vs excess
# crit.lapses <- sleep.nonap.lapses$coefficients$random$subj + sleep.nonap.lapses$coefficients$fixed[3]
# crit.lapses <- data.frame(crit.lapses)
# 
# crit.lapses = rep(crit.lapses$crit, each = 6)
# data.nonap = cbind(data.nonap, critical.lapses = crit.lapses)
# data.nonap = data.nonap %>% mutate(excess.lapses = TWT - critical.lapses*day_num)
# 
# data.nonap = data.nonap %>% mutate(excess.lapses = ifelse(excess.lapses < 0, 0, excess.lapses))
# 
# ggplot(data.nonap, aes(x = excess.lapses, y = lapses)) +
#   stat_summary(fun = mean, geom = "point", aes(group = as.factor(condition), color = as.factor(condition)), alpha = 0.3) +
#  # geom_smooth(color = "black") +
#   stat_summary(fun = mean, geom = "line", aes(y = plapses), size = 1, alpha = 0.5) +
#   #geom_line(aes(y = pred, group = subj)) +
#   labs(x = "excess wakefulness", y = "mean lapses", title = "Mean Lapses vs Excess Wakefulness", color = "condition")

#lapses predictions
data.nonap$plapses = predict(TST.nonap.lapses, data.nonap)
data.nonap$plapses[is.nan(data.nonap$plapses)] <- 0.0

#graphing lapses vs excess
crit.lapses <- TST.nonap.lapses$coefficients$random$subj + TST.nonap.lapses$coefficients$fixed[3]
crit.lapses <- data.frame(crit.lapses)

#crit.lapses %>% filter(crit < 15)
#quantile(crit.lapses$crit, seq(0, 1, 0.10))

# crit.lapses = crit.lapses %>% 
#   mutate(z_latency = as.vector(scale(crit)))
# crit.lapses = crit.lapses %>%  filter(z_latency < -2)

crit.names <- data.frame(subj = rownames(crit.lapses))
crit.lapses$subj <- crit.names$subj

subj.irreg <- data.nonap %>% group_by(subj) %>% count() %>% filter(n < 6) %>% select(subj)
data.nonap <- data.nonap %>% filter(!subj %in% c("NFS2073", "NFS4042", "NFS5063"))

crit.lapses <- crit.lapses %>% filter(!subj %in% c("NFS2073", "NFS4042", "NFS5063"))

#filtering to match participants in both datasets

subj.nonap <-  data.nonap %>% group_by(subj) %>% count() %>% select(subj)
crit.lapses <- crit.lapses %>% filter(subj %in% subj.nonap$subj)
subj.crit <-  crit.lapses %>% group_by(subj) %>% count() %>% select(subj)
data.nonap <- data.nonap %>% filter(subj %in% subj.crit$subj)

#
#nrow(data.nonap)/6
#nrow(crit.lapses)

crit.lapses = rep(crit.lapses$crit, each = 6)
data.nonap = cbind(data.nonap, critical.lapses = crit.lapses)
data.nonap = data.nonap %>% mutate(excess.lapses = TWT_tst - critical.lapses*day_num)

data.nonap = data.nonap %>% mutate(excess.lapses = ifelse(excess.lapses < 0, 0, excess.lapses))

ggplot(data.nonap, aes(x = excess.lapses, y = lapses)) +
  stat_summary(fun = mean, geom = "point", aes(group = as.factor(condition), color = as.factor(condition)), alpha = 0.3) +
 # geom_smooth(color = "black") +
  stat_summary(fun = mean, geom = "line", aes(y = plapses), size = 1, alpha = 0.5) +
  #geom_line(aes(y = pred, group = subj)) +
  labs(x = "excess wakefulness", y = "mean lapses", title = "Mean Lapses vs Excess Wakefulness", color = "condition")

```

To conclude, the estimate of critical waking duration (17.13) was greater than that of Van Dongen's estimate (15.84). Our results seem to suggest that for the average adolescent (based on the sample used for this study), 6.87h of sleep would be sufficient to prevent the build of neurobehavioral deficits at least in the context of the PVT task.